\emph{On Logics Extended with Embedding-Closed Quantifiers}

### 2020-11-14 paper.tex
Generalize the finite case definability theorem

\begin{theorem}\label{log_eq}
Let $\tau$ be a finite relational vocabulary and $\mathcal{H}$ the class of all
finite homogeneous $\tau$-structures. Suppose that $\mathcal{Q}$ is a finite set
of embedding-closed quantifiers of finite width, and $S \subseteq \mathcal{H}$
is a class that is definable over $\mathcal{H}$ by an
$\mathcal{L}^{\omega}_{\infty\omega}(\mathcal{Q})$-sentence $\varphi$.
Then $S$ is nearly definable by an existential $\mathcal{L}_{\omega\omega}$-sentence.
\end{theorem}
\begin{proof}
Let $\mathcal{H}' = \mathcal{H} \cap V_{\omega}$ where $V_{\omega}$ is the
$\omega$-th level in the cumulative hierarchy of sets.
Then, since $V_{\omega}$ is the set of all hereditarily finite sets,
$\mathcal{H}'$ is a set such that for every $\mathfrak{A} \in \mathcal{H}$ there
is $\mathfrak{A}' \in \mathcal{H}'$ isomorphic to $\mathfrak{A}$ so by Lemma
\ref{stabil} there is a choice function
$F\colon \mathcal{P}(\mathcal{H}') \rightarrow \mathcal{H}'$ such that
for all $T \subseteq \mathcal{H}'$, $S \ne \emptyset$, $F(T)$ stabilizes
$\varphi$ in $T$ and $F(\emptyset) = \emptyset$.
Now for all $i < \omega$ we define sets $\mathcal{H}_i \subseteq \mathcal{H}$
and $\tau$-structures $\mathfrak{A}_i$ by setting $\mathcal{H}_0 = \mathcal{H}'$,
$\mathfrak{A}_i = F(\mathcal{H}_i)$ and
\[
	\mathcal{H}_{i+1} = \mathcal{H}_i \setminus \{ \mathfrak{B} \in \mathcal{H} \colon 
	\mathfrak{B} \leq \mathfrak{A}_i \text{ or } \mathfrak{A}_i \leq \mathfrak{B} \}.
\]
Note that $\mathcal{H}_i \subseteq \mathcal{H}_j$ when $i \geq j$.
Let $I = \{i<\omega \colon \mathfrak{A}_i \ne \emptyset \}$.
We claim that for all $i,j \in I$, $i \ne j$, it is the case that
$\mathfrak{A}_i$ and $\mathfrak{A}_j$ are incomparable.
Let $i,j \in I$ and $i < j$.
Assume to the contrary that $\mathfrak{A}_i \leq \mathfrak{A}_j$ or
$\mathfrak{A}_j \leq \mathfrak{A}_i$.
We have $\mathfrak{A}_i \in \mathcal{H}_i$ so from the definition of
$\mathcal{H}_{i+1}$ it follows that $\mathfrak{A}_j \notin \mathcal{H}_{i+1}$.
Now since $\mathcal{H}_j \subseteq \mathcal{H}_{i+1}$ we have
$\mathfrak{A}_j \notin \mathcal{H}_j$, contradiction.
Thus, $\{\mathfrak{A}_i\colon i\in I \}$ is an antichain whose cardinality is
finite by Lemma \ref{shrinking}, and we have
$\mathcal{H} = \mathcal{G}_0 \cup \mathcal{G}_1$ where
\[
	\mathcal{G}_0 = \{\mathfrak{B}\in\mathcal{H} \colon 
	\mathfrak{B} \lneq \mathfrak{A}_i \text{ for some } i \in I \}
\]
is finite up to isomorphism,
and for all $\mathfrak{B} \in \mathcal{G}_1$,
if $\mathfrak{A}_i\leq\mathfrak{B}$, then both $\mathfrak{A}_i$ and
$\mathfrak{B}$ agree on $\varphi$ because $\mathfrak{A}_i$ stabilizes
$\varphi$ in $\mathcal{H}$.
Thus it is easy to see that for each
$\mathfrak{A}\in\mathcal{H} \setminus \mathcal{G}_0$ we have
$\mathfrak{A} \vDash \varphi \Leftrightarrow \mathfrak{A} \vDash \psi$ where
\[
\psi := \bigvee_{\substack{i\in I,\\\mathfrak{A}_i\vDash\varphi}}\vartheta_i,
\]
and all $\vartheta_i$ are existential $\mathcal{L}_{\omega \omega}$-sentences
for which $\mathfrak{A}_i \vDash \vartheta_i$ if and only if
$\mathfrak{A}_i \leq \mathfrak{A}$.

We claim that $\psi$ defines $S \setminus \mathcal{G}_0$ over $\mathcal{H}$.
To show this, let $\mathfrak{A} \in \mathcal{H}$ be such that
$\mathfrak{A} \vDash \psi$.
Then for some $\mathfrak{A}_i$ we have $\mathfrak{A}_i \leq \mathfrak{A}$ and
$\mathfrak{A}_i \vDash \varphi$.
Furthermore, since $\mathfrak{A}_i$ stabilizes $\varphi$ in $\mathcal{H}$, we
have $\mathfrak{A} \vDash \varphi$, and therefore $\mathfrak{A} \in S$.
In addition, it is easy to see that $\mathfrak{A} \notin \mathcal{G}_0$ so
$\mathfrak{A} \in S \setminus \mathcal{G}_0$.
For the other direction, if $\mathfrak{A} \in S \setminus \mathcal{G}_0$,
then $\mathfrak{A} \vDash \varphi$ and, by our earlier observation, we have
$\mathfrak{A} \vDash \psi$.
Thus, since $\mathcal{G}_0$ is finite up to isomorphism, the class $S$ is nearly
definable by an existential $\mathcal{L}_{\omega\omega}$-sentence.
\end{proof}

\begin{corollary}
Let $\tau$ be a finite relational vocabulary and $\mathcal{H}$ the class of all
finite homogeneous $\tau$-structures.
Suppose that $\mathcal{Q}$ is a finite set of embedding-closed quantifiers of
finite width.
Then every $\mathcal{L}^{\omega}_{\infty\omega}(\mathcal{Q})$-sentence is
definable over $\mathcal{H}$ by a $\Sigma^0_2$ sentence of
$\mathcal{L}_{\omega\omega}$.
\end{corollary}
\begin{proof}
Let $\varphi \in \mathcal{L}^{\omega}_{\infty\omega}(\mathcal{Q})[\tau]$.
By Theorem \ref{log_eq}, there is an existential
$\mathcal{L}_{\omega\omega}$-sentence $\psi$ and a class
$S \subseteq \mathcal{H}$ such that $S$ is finite up to isomorphism and for all
$\mathfrak{A} \in \mathcal{H}$ we have
\[
	\mathfrak{A} \vDash \psi 
	\Leftrightarrow \mathfrak{A} \vDash \varphi \text{ and } \mathfrak{A} \notin S,
\]
from which it follows that for all $\mathfrak{A} \in \mathcal{H}$ we have
\[
	\mathfrak{A} \vDash \psi \vee \bigvee_{\mathfrak{B} \in S} \chi_{\mathfrak{B}} 
	\Leftrightarrow \mathfrak{A} \vDash \varphi,
\]
where each $\chi_{\mathfrak{B}}$ is a $\Sigma_2^0$ sentence of
$\mathcal{L}_{\omega \omega}$ that describes the isomorphism type of $\mathfrak{B}$.
\end{proof}

>>>>

\begin{theorem}\label{log_eq}
Let $\tau$ and $\sigma$ be finite relational vocabularies,
$(\Psi, (\psi_R)_{R \in \sigma})$ an 
$\mathcal{L}_{\infty \omega}(\mathcal{Q}_{\emb})$-interpretation of $\sigma$ in
$\tau$ and
\[
\mathcal{H} = \{\Psi(\mathfrak{A}) 
\colon \mathfrak{A} \text{ is a finite homogeneous } \tau \text{-structure}\}.
\]
Suppose that $\mathcal{Q}$ is a finite set
of embedding-closed quantifiers of finite width, 
and $\mathcal{K} \subseteq \mathcal{H}$
is a class that is definable over $\mathcal{H}$ by an
$\mathcal{L}^{\omega}_{\infty\omega}(\mathcal{Q})$-sentence $\varphi$.
Then $\mathcal{K}$ is nearly definable by an existential 
$\mathcal{L}_{\omega\omega}$-sentence.
\end{theorem}
\begin{proof}
Let $\mathcal{H}' = \mathcal{H} \cap V_{\omega}$ where $V_{\omega}$ is the
$\omega$-th level in the cumulative hierarchy of sets.
Then, since $V_{\omega}$ is the set of all hereditarily finite sets,
$\mathcal{H}'$ is a set such that for every $\mathfrak{A} \in \mathcal{H}$ there
is $\mathfrak{A}' \in \mathcal{H}'$ isomorphic to $\mathfrak{A}$ so by Lemma
\ref{stabil} there is a choice function
$F\colon \mathcal{P}(\mathcal{H}') \rightarrow \mathcal{H}'$ such that
for all $S \subseteq \mathcal{H}'$, $S \ne \emptyset$, $F(S)$ stabilizes
$\varphi$ in $T$ and $F(\emptyset) = \emptyset$.
Now for all $i < \omega$ we define sets $\mathcal{H}_i \subseteq \mathcal{H}$
and $\tau$-structures $\mathfrak{A}_i$ by setting $\mathcal{H}_0 = \mathcal{H}'$,
$\mathfrak{A}_i = F(\mathcal{H}_i)$ and
\[
	\mathcal{H}_{i+1} = \mathcal{H}_i \setminus \{ \mathfrak{B} \in \mathcal{H} \colon 
	\mathfrak{B} \leq \mathfrak{A}_i \text{ or } \mathfrak{A}_i \leq \mathfrak{B} \}.
\]
Note that $\mathcal{H}_i \subseteq \mathcal{H}_j$ when $i \geq j$.
Let $I = \{i<\omega \colon \mathfrak{A}_i \ne \emptyset \}$.
We claim that for all $i,j \in I$, $i \ne j$, it is the case that
$\mathfrak{A}_i$ and $\mathfrak{A}_j$ are incomparable.
Let $i,j \in I$ and $i < j$.
Assume to the contrary that $\mathfrak{A}_i \leq \mathfrak{A}_j$ or
$\mathfrak{A}_j \leq \mathfrak{A}_i$.
We have $\mathfrak{A}_i \in \mathcal{H}_i$ so from the definition of
$\mathcal{H}_{i+1}$ it follows that $\mathfrak{A}_j \notin \mathcal{H}_{i+1}$.
Now since $\mathcal{H}_j \subseteq \mathcal{H}_{i+1}$ we have
$\mathfrak{A}_j \notin \mathcal{H}_j$, contradiction.
Thus, $\{\mathfrak{A}_i\colon i\in I \}$ is an antichain whose cardinality is
finite by Lemma \ref{shrinking}, and we have
$\mathcal{H} = \mathcal{G}_0 \cup \mathcal{G}_1$ where
\[
	\mathcal{G}_0 = \{\mathfrak{B}\in\mathcal{H} \colon 
	\mathfrak{B} \lneq \mathfrak{A}_i \text{ for some } i \in I \}
\]
is finite up to isomorphism,
and for all $\mathfrak{B} \in \mathcal{G}_1$,
if $\mathfrak{A}_i\leq\mathfrak{B}$, then both $\mathfrak{A}_i$ and
$\mathfrak{B}$ agree on $\varphi$ because $\mathfrak{A}_i$ stabilizes
$\varphi$ in $\mathcal{H}$.
Thus it is easy to see that for each
$\mathfrak{A}\in\mathcal{H} \setminus \mathcal{G}_0$ we have
$\mathfrak{A} \vDash \varphi \Leftrightarrow \mathfrak{A} \vDash \psi$ where
\[
\psi := \bigvee_{\substack{i\in I,\\\mathfrak{A}_i\vDash\varphi}}\vartheta_i,
\]
and all $\vartheta_i$ are existential $\mathcal{L}_{\omega \omega}$-sentences
for which $\mathfrak{A}_i \vDash \vartheta_i$ if and only if
$\mathfrak{A}_i \leq \mathfrak{A}$.

We claim that $\psi$ defines $\mathcal{K} \setminus \mathcal{G}_0$ over 
$\mathcal{H}$.
To show this, let $\mathfrak{A} \in \mathcal{H}$ be such that
$\mathfrak{A} \vDash \psi$.
Then for some $\mathfrak{A}_i$ we have $\mathfrak{A}_i \leq \mathfrak{A}$ and
$\mathfrak{A}_i \vDash \varphi$.
Furthermore, since $\mathfrak{A}_i$ stabilizes $\varphi$ in $\mathcal{H}$, we
have $\mathfrak{A} \vDash \varphi$, and therefore 
$\mathfrak{A} \in \mathcal{K}$.
In addition, it is easy to see that $\mathfrak{A} \notin \mathcal{G}_0$ so
$\mathfrak{A} \in \mathcal{K} \setminus \mathcal{G}_0$.
For the other direction, if 
$\mathfrak{A} \in \mathcal{K} \setminus \mathcal{G}_0$,
then $\mathfrak{A} \vDash \varphi$ and, by our earlier observation, we have
$\mathfrak{A} \vDash \psi$.
Thus, since $\mathcal{G}_0$ is finite up to isomorphism, the class $\mathcal{K}$ is nearly
definable by an existential $\mathcal{L}_{\omega\omega}$-sentence.
\end{proof}

\begin{corollary}
Let $\tau$ and $\sigma$ be finite relational vocabularies,
$(\Psi, (\psi_R)_{R \in \sigma})$ an 
$\mathcal{L}_{\infty \omega}(\mathcal{Q}_{\emb})$-interpretation of $\sigma$ in
$\tau$ and
\[
\mathcal{H} = \{\Psi(\mathfrak{A}) 
\colon \mathfrak{A} \text{ is a finite homogeneous } \tau \text{-structure}\}.
\]
Suppose that $\mathcal{Q}$ is a finite set of embedding-closed quantifiers of
finite width.
Then every $\mathcal{L}^{\omega}_{\infty\omega}(\mathcal{Q})$-sentence is
definable over $\mathcal{H}$ by a $\Sigma^0_2$ sentence of
$\mathcal{L}_{\omega\omega}$.
\end{corollary}
\begin{proof}
Let $\varphi \in \mathcal{L}^{\omega}_{\infty\omega}(\mathcal{Q})[\sigma]$.
By Theorem \ref{log_eq}, there is an existential
$\mathcal{L}_{\omega\omega}$-sentence $\psi$ and a class
$S \subseteq \mathcal{H}$ such that $S$ is finite up to isomorphism and for all
$\mathfrak{A} \in \mathcal{H}$ we have
\[
	\mathfrak{A} \vDash \psi 
	\Leftrightarrow \mathfrak{A} \vDash \varphi \text{ and } \mathfrak{A} \notin S,
\]
from which it follows that for all $\mathfrak{A} \in \mathcal{H}$ we have
\[
	\mathfrak{A} \vDash \psi \vee \bigvee_{\mathfrak{B} \in S} \chi_{\mathfrak{B}} 
	\Leftrightarrow \mathfrak{A} \vDash \varphi,
\]
where each $\chi_{\mathfrak{B}}$ is a $\Sigma_2^0$ sentence of
$\mathcal{L}_{\omega \omega}$ that describes the isomorphism type of $\mathfrak{B}$.
\end{proof}
